\documentclass[a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{times}
\usepackage{float}
\restylefloat{table}
\usepackage{multicol}
\usepackage{wasysym}
\usepackage{fancyvrb}
\fvset{tabsize=3}
\fvset{fontsize=\small}
\usepackage{enumerate}
\newcommand{\code}{\texttt}

\title{\huge{Curriculum Vitae for Mats Rydberg}}
\date{}

\begin{document}

\maketitle

\section*{Introduction}
Hello, and welcome to my CV! My name is Mats Rydberg, author of this document. My intention with this piece is to show you who I am, what I have learned during my career, and my key specialities as an employee. I hope that you will find the information valuable and that my documented skills do well in describing myself as a potential team-member at your company. Enjoy!

\section*{Personal and contact information}
\begin{tabular}{ll}

Full name & Mats Rydberg \\
Social security number & 19880623-8518 \\
Address & Kämnärsvägen 41C \\
& 22646 Lund \\
&  \\
E-mail & \code{rydberg.mats@gmail.com} \\
Telephone number & 076-9226257 \\
LinkedIn profile & \url{se.linkedin.com/pub/mats-rydberg/38/96a/407/} \\

\end{tabular}



\subsection*{Short autobiography}
I was born in 1988 in Umeå, Sweden, the third child to a teacher. At 12 years of age my family moved south to outside Lund in Skåne, where I was brought up until leaving for university studies in Helsingborg in early 2008. Later the same year I moved back into central Lund and started my five-year education programme of Engineering in Computer Science at LTH. The planned examination date was the summer of 2013, but the summer before that, my first child was born. This had the consequence of me going on parental leave during the spring semester of 2013, prolonging my education by one-half of a year. At the moment of writing, my examination is planned for late February 2014.

In 2011, I was hired as a summer employee at a small company called FindAWay, localized in Malmö. My main tasks were to delevop, from scratch, the back-end of a website aimed at providing high-quality search results for potential flight travellers. This included set-up and configuration of a webserver and database, development in Java Enterprise Edition and interface adaption to existing web services providing flight travel data.

During my education, I was employed as a teaching assistant by the department of Computer Science on several courses, including Java and C++ programming. The main work tasks were to obtain a deep understanding of the programming tasks, in order to be able to then assist and grade students of the course (who were assigned to solve said tasks).

\section*{Education}
\begin{itemize}
 \item Elementary school education at Umeå Waldorfskola (5 years) and Rudolf Steinerskolan i Lund (4 years), 1995-2004.
 
 \item High school education (gymnasium) at Rudolf Steinerskolan i Lund, 2004-2007.
 \begin{itemize}
  \item A broad programme, including natural sciences, humanitarian studies and crafting. Favourite subjects were math, history and sport.
 \end{itemize}
 
 \item History of Ideas at Lund University, 30 credits, fall 2007.
 \begin{itemize}
  \item Examination with top grade of 'VG'. Main interest was history of philosophy.
 \end{itemize}
 
 \item Pre-University course in Technical Sciences at the Faculty of Engineering (LTH) at Lund University, 30 credits, spring 2008.
 \begin{itemize}
  \item Extension of high school programme with deeper studies in math and physics.
 \end{itemize}
 
 \item Master of Science in Engineering, Computer Science and Engineering ('Civil Engineer') at the Faculty of Engineering (LTH) at Lund University, 300 credits, fall 2008-present.
 \begin{itemize}
  \item Central themes in this programme are math, programming and software development. Elementary programming is taught in Java, using Eclipse IDE. Much of the course literature is written in English, and some courses are given in English. Many software development courses include working in teams and/or pairs. Last two years are focused into one specialization tree. The final semester is an individual project, my subject being algorithms.
  \item I have completed courses worth 270 credits. My average grade is 4.3 (on a scale of 3/4/5, where 3 is passed and 5 is top grade).
 \end{itemize}

\end{itemize}

\subsection*{Specialization and key skills}
My specialization during my Computer Science studies was software development, with a focus on programming and algorithms. Key courses were C++ Programming, Advanced Computer Security, Advanced Algorithms, Functional Programming, Configuration Management and Optimising Compilers.

My individual project, carried out fall 2013 (present), revolve around algorithms. My main task has been to implement a state-of-the-art algorithm for graph colouring (more precisely, for computing the chromatic polynomial of a graph), to perform automated simulations on the resulting program, to compare my experimental results to previously 'best possible'-achieving software, and to document my results in a scientific report. While my work has in large been research, the pure programmatic tasks have been taughtful. I have worked in a C/C++ environment, using the interfaces and reading the documentation of third-party libraries. My program evolved from single-threaded to parallelized. I have dealt with large memory requirements (problem sizes in the field of graph colouring increases exponentially), and the issues that come with it.

I write and speak fluent Swedish (mother tongue) and English. From language similarity, I can communicate in Danish and Norwegian. I am educated in German, and have elementary knowledge of Spanish.

\section*{Professional experience}

\subsection*{Developer at FindAWay}
During the summer of 2011, I spent six weeks at FindAWay Technology AB, a small Malmö-based company with the aims of a brand new website for flight travel search. I developed the back-end of this system, including setting up and configuring an Apache TomCat web server, a MySQL database with a Java Enterprise Edition driver connection, and configuring all of it to work inside a NetBeans Integrated Development Environment (IDE). Implementation of the system was done in minimal SQL, with pure Java EE (version 6) as the focus. The work included adaption to interfaces of existing web services for flight travel search via XML.

%The website was never launched, but the business ideas were later bought up by Momondo, at which the software is now integrated with their systems.

Reference: Lars Kristensson, \code{lars@k-konsulting.se}.

\subsection*{Teaching Assistant at LTH}
In parallel with my studies at LTH, I have been employed as a teaching assistant during programming exercises in the following courses:
\begin{itemize}
 \item Programming in Java (basic level).
 \begin{itemize}
  \item I have taught this course twice.
 \end{itemize}

 \item Programming in Java (intermediate level).
 \item C++-programming (intermediate level).
 \begin{itemize}
  \item I have also been involved in a re-design process of the programming tasks of this course.
 \end{itemize}

\end{itemize}
The work has deepened my understanding of programming as a whole, in the programming languages in particular, and improved my interaction with people in a professional manner. My main tasks have been to thoroughly learn certain exercises, in order to aid others in understanding the principles which lie behind them. As a teaching assistant, I was also responsible for grading the students (on a passed/failed scale), which included two key skills: doing quality assurance to ensure the work performed met the specified requirements of the given task, and obtaining a personal, but professional, contact with the students to get a picture of the knowledge attained in each individual.

\subsection*{Voluntary projects}
During my time as a student, I've been engaged in a number of voluntary projects. Two deserve special mention.

I spent one year as the cheif responsible of quality assurance and general student's counsel for my education programme, a position which is part of the student union. The position is meriting in terms of ability to take individual responsibility, to start and perform your own projects, and to communicate with people.

I was one of the (two) communication and media contact responsible people for the project Rosa Marschen, a charity fund for breast cancer research. The concept is a march through the city of Lund, with attendance fees going straight to the fund. 2013 was the first year, and we succeeded in drawing near 300 people in our march on October 5th. My job was to communicate with newspapers and other media, and I figured in a few articles online and on paper. The position was meriting in terms of professional communication ability, and ability to work as a team in our project group.

\end{document}
